Make any stupid plot you want

### 2021-02-24 draft/sections/intro.tex
gog rework

\section{Background}
Visual algorithms that display information are designed such that the structure of data is assumed, as described in Tory and Möller's taxonomy \cite{toryRethinkingVisualizationHighLevel2004}. Specifically they note that discrete and continuous data and their attributes form a discipline independent design space \cite{pousmanCasualInformation2007}. This assumption is evident in the information visualization software design patterns categorized by Heer and Agrawala\cite{HeerSoftware2006} and can often be found in modern general purpose visualization libraries.  In proposing a new architecture, we contrast the trade offs libraries make, describe different types of data continuity, and discuss metrics by which a visualization library is traditionally evaluated.

\subsection{Tools}
\label{sec:intro_data_tools}
The library driving this rearchitecture, Matplotlib, aims to be flexible enough that developers can build domain specific libraries on top of it. To preserve this flexibility, Matplotlib enforces minimal constraints on what sorts of data the user is allowed to input and supports very many visual algorithms, from primative marks to computationally complex visualizations. Instead of enforcing constraints at the API level, Matplotlib carries implicit assumptions about data continuity in how each function interfaces with the input data. This has lead to poor API consistency and brittle code as every visual algorithm has a very different point of view on how the data is structured.

A commonly cited alternative is the family of tools built on top of Wilkenson's Grammar of Graphics (GoG) \cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}. This framework is very popular in the data visualization community because it lends itself to a declarative interface \cite{heerDeclarative2010} which allows end users to describe the visualization they are trying to create.  Grammar oriented approaches describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building the elements.

@@

\subsection{Data}
\label{sec:intro_data}
As mentioned, one of the drivers of this work was to facilitate building libraries that could natively support domain specific data containers. Fiber bundles are one such way model to model containers, as Butler proposes because they encode the continuity of the data separately from the types of variables and are flexible enough to support discrete and ND continuous datasets \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}. Since Butler's model lacks a robust way of describing the variables, we fold in Spivak's Simplacial formulation of databases \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES} so that we can encode a schema like description of the data in the fiber bundle.

>>>>

\section{Background}
One of the reasons we developed a new formalism rather than adopting the architecture of an existing library is that most information visualization software design patterns, as categorized by Heer and Agrawala\cite{HeerSoftware2006}, are tuned to very specific data structures. This in turn restricts the design space of visual algorithms that display information (the visualization types the library supports) since the algorithms are designed such that the structure of data is assumed, as described in Tory and Möller's taxonomy \cite{toryRethinkingVisualizationHighLevel2004}. In proposing a new architecture, we contrast the trade offs libraries make, describe different types of data continuity, and discuss metrics by which a visualization library is traditionally evaluated.

\subsection{Tools}
\label{sec:intro_data_tools}
The library driving this rearchitecture, Matplotlib, aims to be flexible enough that developers can build domain specific libraries on top of it. To preserve this flexibility, Matplotlib enforces minimal constraints on the types of data the user is allowed to input and supports very many types of graphics, from primitive marks to computationally complex visualizations. Instead of enforcing constraints at the API level, Matplotlib carries implicit assumptions about data continuity in how each function interfaces with the input data. This has lead to poor API consistency and brittle code as every visual algorithm has a very different point of view on how the data is structured.

A commonly cited alternative is the family of tools built on top of Wilkenson's Grammar of Graphics (GoG) \cite{wilkinsonGrammarGraphics2005}, including ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, protovis\cite{bostockProtoviz2009} and D3 \cite{bostockDataDrivenDocuments2011}, vega\cite{satyanarayanDeclarativeInteractionDesign2014} and altair\cite{vanderplasAltairInteractiveStatistical2018}. The GoG family of tools are very strongly built on the assumption that data is a relational framework; this is explicit in Mackinlay's A Presentation Tool (APT)\cite{mackinlayAUTOMATICDESIGNGRAPHICAL1987} formalism on which GoG is build. This framework is very popular in the data visualization community because it lends itself to a declarative interface \cite{heerDeclarative2010} which allows end users to describe the visualization they are trying to create. Grammar oriented approaches describe how to compose visual elements into a graphical design \cite{wongsuphasawatNavigatingWideWorld2021}, while we are proposing a framework for building the elements. A clear example of this distinction is that the GoG grammar includes computation and aggregation of the table as part of the grammar, while we propose that computations are specific to domains and only try to describe them when they are specifically part of the visual encoding - for example mapping data to a color.

assumptions on each specific component of the visualization, which in turn allows for building composable

@@

\subsection{Data}
\label{sec:intro_data}
Discrete and continuous data and their attributes form a discipline independent design space \cite{pousmanCasualInformation2007}, so one of the drivers of this work was to facilitate building libraries that could natively support domain specific data containers that do not make assumptions about data continuity. Fiber bundles are one such way model to model containers, as Butler proposes because they encode the continuity of the data separately from the types of variables and are flexible enough to support discrete and ND continuous datasets \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}. Since Butler's model lacks a robust way of describing the variables, we fold in Spivak's Simplacial formulation of databases \cite{spivakDatabasesAreCategories2010,spivakSIMPLICIALDATABASES} so that we can encode a schema like description of the data in the fiber bundle.
